---
title: "Data Wrangling with R"
subtitle: "IBM 320 Section 02"
author: "Elijah Kates"
date: "today"
format: 
    html:
        theme: cosmo
        toc: true
        toc-depth: 3
        toc-location: left
        embed-resources: true
execute: 
  freeze: auto
---

::: {callout-note}
Note: This codebook will follow the structure and content of the published codebook of the project:
[Data Wrangling with Tidyverse](https://jmjung.quarto.pub/m05-1-principles-data-wrangling-with-tidyverse-in-r/)
:::

# Overview

## Learning Outcomes

By the end of this module, you will be able to:

-   Describe the concept of Data Wrangling
-   Describe how Tibbles are different from data frames
-   Explain how to convert wide or long data to tidy data
-   Explain how to merge relational data sets using join functions (next module)
-   Explain how to use grouped mutates and filter together
-   Be familiar with major dplyr functions for transforming data
-   Create a new variable with `mutate()` and `case_when()`
-   Use the pipe operator to shape data for analysis and visualization

## The textbook chapters to cover

-   Ch3: Data Transformation\
-   Ch5: Data Tidying\
-   Ch13: Numbers

# Introduction to Data Wrangling

Data wrangling refers to the process of cleaning, transforming, and organizing raw data into a format that can be used for analysis.

# Loading the Library

```{r}
library(tidyverse)
```

# What is Tidyverse

The **tidyverse** is a collection of R packages designed for data science that share a common philosophy and grammar.

Core packages include:

-   **ggplot2** for data visualization\
-   **dplyr** for data manipulation\
-   **tidyr** for data tidying\
-   **readr** for data import\
-   **purrr** for functional programming\
-   **tibble** for modern data frames\
-   **stringr** for string manipulation\
-   **forcats** for categorical variables

# Initial Data Preparation and Exploration

```{r}
library(nycflights13)
data()
```

# Data Wrangling in dplyr

The **dplyr** package provides functions for manipulating data.

Note: use `Shift + Option + I` for multi-cursor activation

`airlines` Airline names.\
`airports` Airport metadata.\
`flights` Flights data.\
`planes` Plane metadata.\
`weather` Hourly weather data.

```{r}
flights <- flights

flights |> 
    count(year, month) |> 
    arrange(desc(n)) |> 
    mutate(month = as_factor(month)) |> 
    mutate(month = fct_reorder(month, n)) |> 
    ggplot(aes(month, n, fill = year)) + 
    geom_col(show.legend = FALSE) + coord_flip() + 
    labs(x = "Month", y = "# of Flights", title = "# of Flights by Month during Year 2013 at New York"
    )
```
